---
title: "`SpatialData`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('SpatialData')`"
author: 
  - name: Helena Lucia Crowell
  - name: Louise Deconinck
  - name: Artür Manukyan
  - name: Dario Righelli
  - name: Estella Dong
  - name: Vince Carey
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{SpatialData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

```{r load-libs}
library(ggplot2)
library(patchwork)
library(ggnewscale)
library(SpatialData)
```

# Introduction

The `SpatialData` package contains a set of reader and plotting functions for 
spatial omics data stored as [SpatialData](https://spatialdata.scverse.org/en/latest/index.html) 
`.zarr` files, following [OME-NGFF specs](https://ngff.openmicroscopy.org/latest/#image-layout).

Each `SpatialData` object is composed of five layers: images, labels, shapes, 
points, and tables. Each layer may contain an arbitrary number of elements.

Images and labels are represented as `ZarrArray`s (`r BiocStyle::Biocpkg("Rarr")`). 
Points and shapes are represented as `r BiocStyle::CRANpkg("arrow")` objects linked 
to an on-disk *.parquet* file. As such, all data are represented out of memory.

Element annotation as well as cross-layer summarizations (e.g., count matrices)
are represented as `r BiocStyle::Biocpkg("SingleCellExperiment")` as tables.

## Handling

For demonstration, we read in a toy dataset that is available through the package:

```{r blobs-read}
x <- file.path("extdata", "blobs.zarr")
x <- system.file(x, package="SpatialData")
(x <- readSpatialData(x))
```

`SpatialData` object behave like a list, thereby supporting flexible accession, 
e.g., via `$` and `[[`, using character or integer indices etc. Specifically,

- `image/label/shape/point/table()` retrieve one element
- `images/labels/shapes/points/tables()` retrieve one layer
- `image/label/shape/point/tableNames()` retrieve element names

Let's demonstrate these capabilities using the image layer:

```{r get, results="hide"}
shapeNames(x) # there are two images
shapes(x) # this is a list of them
shape(x, i=1) # this is the 1st one
shape(x, "blobs_polygons") # this is the 3nd one
x$shapes$blobs_circles # list-like handling also works
```

Each element is made up of two key components: 

- `data`: the actual array or table (for images/labels and shapes/points, respectively)
- `meta`: list of metadata retrieved from an OME-NGFF compliant *.zattrs* (stored as `Zattrs`)

```{r data/meta}
data(shape(x))
meta(shape(x)) 
```

## Multiscales

```{r ms}
image(x, "blobs_multiscale_image")
```

## Coordinates

`Zattrs` specify an explicit relationship between elements and coordinate systems.
We can represent these are a graph as follows:

```{r cs-graph}
(g <- SpatialData:::.coord2graph(x))
graph::plot(g)
```

The above representation greatly facilitates queries of the transformation(s) 
required to spatially align elements. `blobs_labels`, for example, requires a
sequential transformation (scaling and translation) for the `sequence` space:

```{r cs-path}
SpatialData:::.get_path(g, "blobs_labels", "sequence")
```

## Visualization

```{r blobs-plot}
plotSpatialData() +
    plotLabel(x) +
    plotPoint(x, c="genes")
```

# Technologies

## MERFISH

```{r merfish-read}
dir.create(td <- tempfile())
pa <- unzip_merfish_demo(dest=td)
x <- readSpatialData(pa)
```

In this example data, we do not have a `label` for the `shape` polygons. Such labels could be morphological regions annotated by pathologists.

```{r merfish-plot}
i <- sample(length(p <- point(x, i=1)), 2e3)
point(x, "2k") <- p[i]
plotSpatialData() +
    plotPoint(x, i="2k", c="cell_type", s=0.2) +
    new_scale_color() +
    plotShape(x, i="anatomical") +
    scale_color_manual(values=hcl.colors(6, "Spectral")) 
```

We could also show plot the transcript without cell segmentation. 
There are only 2,000 cells, but 3,714,642 molecules, so it takes a long time to compile.

```{r eval=FALSE}
plotSpatialData() +
    plotShape(x, i=2, c="black", f=NA) +
    scale_size_continuous(range=c(0, 3))
```

# MibiTOF

Colorectal carcinoma, 25 MB; no shapes, no points.

```{r mibitof-read}
dir.create(td <- tempfile())
pa <- unzip_spd_demo(zipname="mibitof.zip", dest=td, source="biocOSN")
(x <- readSpatialData(pa))
```

## label

Explore the label related functionalities. In this MibiTOF example, we only have two slots named `"single_molecule"` and `"2k"`.

```{r eval=FALSE}
# object should behave like a names list
x$labels$point16_image
x$labels$point23_image
x$labels$point8_image

# character vector
pointNames(x) 
# list of elements in layer
points(x) 
# single element
point(x, i=1) # default
point(x, i="2k")
# get (meta)data from an element
# (arrow for point/shape, ZarrArray for label/image)
data(point(x)) 
# .zattrs, currently just rendered as 'fromJSON' would
meta(point(x)) 
```

```{r mibitof-plot}
lapply(seq(3), \(i) {
    plotSpatialData() + 
        plotImage(x, i) + 
        new_scale_fill()
}) |> wrap_plots(nrow=1) + plot_layout(guides="collect")
```

# VisiumHD

Mouse intestine, 1GB; 4 image resolutions and 3 shapes at 2, 8, and 16 $\mu$m.

```{r visiumhd-read}
dir.create(td <- tempfile())
nm <- "visium_hd_3.0.0_io.zip"
pa <- unzip_spd_demo(zipname=nm, dest=td, source="biocOSN")
(x <- readSpatialData(pa, tables=NULL))
```

```{r visiumhd-plot}
plotSpatialData() + plotImage(x, i=4)
```

## table

Explore the label related functionalities. In this MibiTOF example, we only have two slots named `"single_molecule"` and `"2k"`. 

```{r eval=FALSE}
# object should behave like a names list
x$table$square_002um
x$table$square_008um
x$table$square_016um

# character vector
tableNames(x) 
# list of elements in layer
tables(x) 
# single element
SpatialData::table(x, i=1) # default
SpatialData::table(x, i="square_002um")
SpatialData::table(x, i="square_008um")
SpatialData::table(x, i="square_016um")
# get (meta)data from an element
# (arrow for point/shape, ZarrArray for label/image)
data(SpatialData::table(x)) 
# .zattrs, currently just rendered as 'fromJSON' would
meta(SpatialData::table(x)) 
```

# CyCIF (MCMICRO output)

Small lung adenocarcinoma, 250 MB; 1 image, 2 labels, 2 tables.

```{r mcmicro-read}
dir.create(td <- tempfile())
nm <- "mcmicro_io.zip"
pa <- unzip_spd_demo(zipname=nm, dest=td, source="biocOSN")
(x <- readSpatialData(pa))
```

# IMC (Steinbock output)

4 different cancers (SCCHN, BCC, NSCLC, CRC), 820 MB; 14 images, 14 labels, 1 table.

```{r steinbock-read}
dir.create(td <- tempfile())
nm <- "steinbock_io.zip"
pa <- unzip_spd_demo(zipname=nm, dest=td, source="biocOSN")
(x <- readSpatialData(pa))
```

# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
