---
title: "Introduction to `SpatialData`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('SpatialData')`"
author: 
  - name: Helena L. Crowell
  - name: Louise Deconinck
  - name: Artür Manukyan
  - name: Dario Righelli
  - name: Estella Dong
  - name: Samuel Gunz
  - name: Yixing E. Dong
  - name: Michael Stadler
  - name: Charlotte Soneson
  - name: Vince Carey
  - name: Hugo Gruson
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{Introduction to SpatialData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

# Introduction

The `r BiocStyle::Biocpkg("SpatialData")` package provides an R interface to the 
[SpatialData](https://spatialdata.scverse.org) framework, a unified ecosystem 
for handling spatial omics data. Developed as part of the `scverse` community, 
`SpatialData` aims to solve the challenges of integrating diverse spatial 
datasets—including imaging, spatial transcriptomics, and proteomics—by 
employing the [OME-NGFF (Next Generation File Format)](https://ngff.openmicroscopy.org) 
standard.

## Data Representation in R

The core of the package is the `SpatialData` class, which serves as a container 
for multiple layers of spatial information:

- **Images**: Multiscale, multi-channel raster data, represented as `ZarrArray` objects via the `r BiocStyle::Biocpkg("Rarr")` and `r BiocStyle::Biocpkg("ZarrArray")` packages.
- **Labels**: Segmentation masks (e.g., cell or nuclei boundaries), also stored as `ZarrArray`s.
- **Points**: High-resolution coordinates (e.g., individual transcript molecules), handled as `r BiocStyle::CRANpkg("arrow")` tables or `r BiocStyle::CRANpkg("duckdb")`-backed data frames for efficient out-of-memory processing.
- **Shapes**: Polygonal regions of interest or circular markers, represented similarly to points but with geometric metadata.
- **Tables**: Functional annotations (e.g., gene expression counts), stored as `r BiocStyle::Biocpkg("SingleCellExperiment")` objects, which link to the spatial elements via `instance_key` and `region` identifiers.

This architecture ensures that `SpatialData` remains scalable, allowing users to 
interact with datasets that exceed available RAM by leveraging on-disk storage 
and lazy loading.

# Getting Started

First, we load the necessary libraries:

```{r load-libs}
library(SpatialData)
library(SingleCellExperiment)
library(ZarrArray)
```

## Reading Data

`SpatialData` objects are typically read from `.zarr` stores. The package provides 
the `readSpatialData()` function to ingest an entire store, as well as specific 
readers for individual elements.

```{r read-data}
# Path to a toy dataset included in the package
path <- system.file("extdata", "blobs.zarr", package="SpatialData")

# Read the complete SpatialData object
sd <- readSpatialData(path)
sd
```

## Accessing Elements

The `SpatialData` object behaves like a named list of layers. You can use standard 
R accessors to retrieve layers and elements:

```{r accessors}
# List names of all image elements
imageNames(sd)

# Retrieve the first label element
lbl <- label(sd, 1)
lbl

# Access the underlying data of a point element
pts <- point(sd, 1)
head(as.data.frame(data(pts)))
```

# Coordinate Systems and Transformations

A key feature of the `SpatialData` framework is its robust handling of 
coordinate systems. Each element can exist in multiple coordinate spaces 
simultaneously, defined by transformations in its metadata.

## Metadata Access

The `meta()` function retrieves the `Zattrs` object, which contains OME-NGFF 
compliant metadata:

```{r metadata}
z <- meta(lbl)
axes(z)
CTname(z)
```

## Applying Transformations

Elements can be transformed between spaces. The `transform()` function resolves 
the necessary steps to project an element into a target coordinate system:

```{r transform}
# Apply a transformation to a point element
pts_transformed <- transform(pts, "global")
```

# Spatial Queries

`SpatialData` supports sophisticated spatial operations such as cropping and 
querying, which are propagated across all layers:

```{r crop}
# Define a bounding box for cropping
box <- list(xmin=10, xmax=50, ymin=10, ymax=50)

# Crop the entire SpatialData object
sd_cropped <- crop(sd, box, j="global")
sd_cropped
```

# Integration with Tables

Tables (SCE objects) are used to annotate spatial elements. They link to specific 
labels or shapes using identifiers:

```{r tables}
# Access the table layer
if (length(tables(sd)) > 0) {
    sce <- table(sd, 1)
    
    # Check which region this table annotates
    region(sce)
    
    # Check the instance key (linking to specific labels/shapes)
    instance_key(sce)
}
```

# Conclusion

The `SpatialData` package brings the power of the `scverse` spatial ecosystem to 
R, providing a scalable and interoperable foundation for spatial omics analysis. 
By adhering to the OME-NGFF standard and integrating with established 
Bioconductor objects like `SingleCellExperiment`, it enables seamless 
workflows for the modern biologist.

# Session Info

```{r sessionInfo}
sessionInfo()
```
