---
title: "`SpatialData`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('SpatialData')`"
author: 
  - name: Helena Lucia Crowell
  - name: Louise Deconinck
  - name: Artür Manukyan
  - name: Dario Righelli
  - name: Estella Dong
  - name: Vince Carey
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{SpatialData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

```{r load-libs}
library(ggplot2)
library(patchwork)
library(ggnewscale)
library(SpatialData)
library(SingleCellExperiment)
```

# Introduction

The `SpatialData` package contains a set of reader and plotting functions for 
spatial omics data stored as [SpatialData](https://spatialdata.scverse.org/en/latest/index.html) 
`.zarr` files that follow [OME-NGFF specs](https://ngff.openmicroscopy.org/latest/#image-layout).

Each `SpatialData` object is composed of five layers: images, labels, shapes, 
points, and tables. Each layer may contain an arbitrary number of elements.

Images and labels are represented as `ZarrArray`s (`r BiocStyle::Biocpkg("Rarr")`). 
Points and shapes are represented as `r BiocStyle::CRANpkg("arrow")` objects linked 
to an on-disk *.parquet* file. As such, all data are represented out of memory.

Element annotation as well as cross-layer summarizations (e.g., count matrices)
are represented as `r BiocStyle::Biocpkg("SingleCellExperiment")` as tables.

## Handling

For demonstration, we read in a toy dataset that is available through the package:

```{r blobs-read}
x <- file.path("extdata", "blobs.zarr")
x <- system.file(x, package="SpatialData")
(x <- readSpatialData(x, anndataR=FALSE))
```

`SpatialData` object behave like a list, thereby supporting flexible accession, 
e.g., via `$` and `[[`, using character or integer indices etc. Specifically,

- `image/label/shape/point/table()` retrieve one element
- `images/labels/shapes/points/tables()` retrieve one layer
- `image/label/shape/point/tableNames()` retrieve element names

Let's demonstrate these capabilities using the image layer:

```{r get, results="hide"}
shapeNames(x) # there are two images
shapes(x) # this is a list of them
shape(x, i=1) # this is the 1st one
shape(x, "blobs_polygons") # this is the 3nd one
x$shapes$blobs_circles # list-like handling also works
```

Each element is made up of two key components: 

- `data`: the actual array or table (for images/labels and shapes/points, respectively)
- `meta`: list of metadata retrieved from an OME-NGFF compliant *.zattrs* (stored as `Zattrs`)

```{r data/meta}
data(shape(x))
meta(shape(x)) 
```

## Annotations

Currently supported methods for the `tables` layers include:

- `getTable()` to retrieve the `table` annotating a given element
- `setTable()` to construct a new annotation for a given element
- `valTable()` to retrieve specific values from a `table`

```{r tbl}
# retrieve 'table' for an element
(t <- getTable(x, i <- "blobs_labels"))
head(colData(t))
meta(t)
# get values from an element's 'table'
valTable(x, i, "channel_0_sum")
# add an artificial 'table' to an element
i <- "blobs_points"
# ...passing a preconstructed 'data.frame'
n <- length(point(x, i))
df <- data.frame(n=runif(n))
y <- setTable(x, i, df)
head(colData(getTable(y, i)))
# ...using a list of data generating functions
f <- list(
    numbers=\(n) runif(n),
    letters=\(n) sample(letters, n, TRUE))
args <- c(list(x, i), f)
y <- do.call(setTable, args)
head(colData(getTable(y, i)))
```

## Transformations

To facilitate .zattrs handling, we provide a set of functions to access and modify its contents. Specifically,

- `axes` to access a given element's dimension names and types
- `CTdata/name/type` to access coordinate transformation components
- `rmv/addCT` to remove, add, or append coordinate transformations

```{r cs-methods}
z <- meta(label(x))
axes(z)
CTdata(z)
CTname(rmvCT(z, "scale"))
CTname(addCT(z, 
    name="D'Artagnan", 
    type="scale", 
    data=c(19, 94)))
```

`Zattrs` specify an explicit relationship between elements and coordinate systems.
We can represent these are a graph as follows:

```{r cs-graph}
(g <- SpatialData:::.coord2graph(x))
graph::plot(g)
```

The above representation greatly facilitates queries of the transformation(s) 
required to spatially align elements. `blobs_labels`, for example, requires a
sequential transformation (scaling and translation) for the `sequence` space:

```{r cs-path}
SpatialData:::.get_path(g, "blobs_labels", "sequence")
```

## Multiscales

`Image/LabelArray`s are linked to potentially multiscale .zarr stores.
Their show method includes the scales available for a given element:

```{r ms-show}
image(x, "blobs_image")
image(x, "blobs_multiscale_image")
```

Internally, multiscale `ZarrArray`s are stored as a list, e.g.:

```{r ms-dims}
i <- image(x, "blobs_multiscale_image")
vapply(i@data, dim, numeric(3))
```

## Visualization

```{r blobs-plot}
# layered
plotSpatialData() +
    plotImage(x) +
    plotLabel(x) +
    plotShape(x, 1) +
    plotShape(x, 3) +
    new_scale_color() +
    plotPoint(x, c="genes")
# multiscales
wrap_plots(nrow=1, lapply(seq(3), \(.) 
    plotSpatialData() + plotImage(x, i=2, k=.)))
```

# Datasets

Data from a variety of technologies has been made available as `SpatialData` .zarr stores here 
[here](https://spatialdata.scverse.org/en/latest/tutorials/notebooks/datasets/README.html).
These, in turn, have been deposited in Bioconductor's NSF Open Storage Network bucket,
and can be retrieved from there with support for caching using `r BiocStyle::Biocpkg("BiocFileCache")`.

We can interrogate the bucket for available (zipped) .zarr archives:

```{r ons-list}
available_spd_zarr_zips()
```

Any of the above can be retrieved (once) into some location, and read into R:

```{r ons-read}
dir.create(td <- tempfile())
pa <- unzip_spd_demo(
    zipname="merfish.zarr.zip", 
    dest=td, source="biocOSN")
(x <- readSpatialData(pa))
```

# Technologies

## MERFISH

In this example data, we do not have a `label` for the `shape` polygons. 
Such labels could be morphological regions annotated by pathologists.

There are only `r length(SpatialData::shape(x, "cells"))` cells, but 
`r format(length(SpatialData::point(x, "single_molecule")), big.mark=",")` 
molecules, so that we downsample a random subset of 2,000 for visualization:

```{r merfish-plot}
i <- sample(length(p <- point(x, i=1)), 2e3)
point(x, "2k") <- p[i]
plotSpatialData() +
    plotImage(x) +
    plotPoint(x, i="2k", c="cell_type", s=0.2) +
    new_scale_color() +
    plotShape(x, i="anatomical") +
    scale_color_manual(values=hcl.colors(6, "Spectral")) 
```

```{r}
qu <- list(xmin=1800, xmax=2400, ymin=5000, ymax=5400)
bb <- geom_rect(do.call(aes, qu), data.frame(qu), col="yellow", fill=NA)
y <- SpatialData(images=list("."=do.call(query, c(list(x=image(x)), qu))))
plotSpatialData() + plotImage(x) + bb | plotSpatialData() + plotImage(y)
```

## MibiTOF

Colorectal carcinoma, 25 MB; no shapes, no points.

```{r mibitof-read}
dir.create(td <- tempfile())
pa <- unzip_spd_demo(
    zipname="mibitof.zip", 
    dest=td, source="biocOSN")
(x <- readSpatialData(pa))
```

```{r mibitof-plot}
wrap_plots(nrow=1, lapply(seq(3), \(.) 
    plotSpatialData() + plotImage(x, .) + plotLabel(x, .)))
```

## VisiumHD

Mouse intestine, 1GB; 4 image resolutions and 3 shapes at 2, 8, and 16 $\mu$m.

```{r visiumhd-read}
dir.create(td <- tempfile())
pa <- unzip_spd_demo(
    zipname="visium_hd_3.0.0_io.zip", 
    dest=td, source="biocOSN")
(x <- readSpatialData(pa, images=4, shapes=3, tables=NULL))
```

```{r visiumhd-plot}
qu <- list(xmin=100, xmax=300, ymin=200, ymax=400)
bb <- geom_rect(do.call(aes, qu), data.frame(qu), col="black", fill=NA)
y <- SpatialData(images=list("."=do.call(query, c(list(x=image(x)), qu))))
plotSpatialData() + plotImage(x) + bb | plotSpatialData() + plotImage(y)
```

## CyCIF (MCMICRO)

Small lung adenocarcinoma, 250 MB; 1 image, 2 labels, 2 tables.

```{r mcmicro-read}
dir.create(td <- tempfile())
pa <- unzip_spd_demo(
    zipname="mcmicro_io.zip", 
    dest=td, source="biocOSN")
(x <- readSpatialData(pa))
```

## IMC (Steinbock)

4 different cancers (SCCHN, BCC, NSCLC, CRC), 820 MB; 14 images, 14 labels, 1 table.

```{r steinbock-read}
dir.create(td <- tempfile())
pa <- unzip_spd_demo(
    zipname="steinbock_io.zip", 
    dest=td, source="biocOSN")
(x <- readSpatialData(pa))
```

# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
