---
title: "`SpatialData`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('SpatialData')`"
author: 
  - name: Helena Lucia Crowell
  - name: Louise Deconinck
  - name: Artür Manukyan
  - name: Dario Righelli
  - name: Estella Dong
  - name: Vince Carey
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{SpatialData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

```{r load-libs}
library(patchwork)
library(SpatialData)
library(SingleCellExperiment)
```

# Introduction

The `SpatialData` package contains a set of reader and plotting functions for 
spatial omics data stored as [SpatialData](https://spatialdata.scverse.org/en/latest/index.html) 
`.zarr` files that follow [OME-NGFF specs](https://ngff.openmicroscopy.org/latest/#image-layout).

Each `SpatialData` object is composed of five layers: images, labels, shapes, 
points, and tables. Each layer may contain an arbitrary number of elements.

Images and labels are represented as `ZarrArray`s (`r BiocStyle::Biocpkg("Rarr")`). 
Points and shapes are represented as `r BiocStyle::CRANpkg("arrow")` objects linked 
to an on-disk *.parquet* file. As such, all data are represented out of memory.

Element annotation as well as cross-layer summarizations (e.g., count matrices)
are represented as `r BiocStyle::Biocpkg("SingleCellExperiment")` as tables.

## Handling

For demonstration, we read in a toy dataset that is available through the package:

```{r blobs-read}
x <- file.path("extdata", "blobs.zarr")
x <- system.file(x, package="SpatialData")
(x <- readSpatialData(x))
```

`SpatialData` object behave like a list, thereby supporting flexible accession, 
e.g., via `$` and `[[`, using character or integer indices etc. Specifically,

- `image/label/shape/point/table()` retrieve one element
- `images/labels/shapes/points/tables()` retrieve one layer
- `image/label/shape/point/tableNames()` retrieve element names

Let's demonstrate these capabilities using the image layer:

```{r get, results="hide"}
names(x@shapes) # there are two shapes
x@shapes # this is a list of them
x@shapes[[1]] # this is the 1st one
x@shapes$blobs_circles # list-like handling also works
```

Each element is made up of two key components: 

- `data`: the actual array or table (for images/labels and shapes/points, respectively)
- `meta`: list of metadata retrieved from an OME-NGFF compliant *.zattrs* (stored as `Zattrs`)

```{r data/meta}
data(x@shapes[[1]])
zattrs(x@shapes[[1]])
```

## Annotations

Currently supported methods for the `tables` layers include:

- `getTable()` to retrieve the `table` annotating a given element
- `setTable()` to construct a new annotation for a given element
- `valTable()` to retrieve specific values from a `table`

TODO...

```{r tbl, eval=FALSE}
# retrieve 'table' for an element
(t <- getTable(x, i <- "blobs_labels"))
head(colData(t))
meta(t)
# get values from an element's 'table'
valTable(x, i, "channel_0_sum")
# add an artificial 'table' to an element
i <- "blobs_points"
# ...passing a preconstructed 'data.frame'
id <- unique(point(x)$instance_id)
df <- data.frame(n=runif(length(id)))
y <- setTable(x, i, df)
head(colData(getTable(y, i)))
# ...using a list of data generating functions
f <- list(
    numbers=\(n) runif(n),
    letters=\(n) sample(letters, n, TRUE))
args <- c(list(x, i), f)
y <- do.call(setTable, args)
head(colData(getTable(y, i)))
```

## Transformations

To facilitate .zattrs handling, we provide a set of functions to access and modify its contents. Specifically,

- `axes` to access a given element's dimension names and types
- `CTdata/name/type` to access coordinate transformation components
- `rmv/addCT` to remove, add, or append coordinate transformations

```{r cs-methods}
z <- zattrs(x@labels$blobs_labels)
ct_axes(z)
ct_name(z)
ct_args(z, "scale")
```

`Zattrs` specify an explicit relationship between elements and coordinate systems.
We can represent these are a graph as follows:

```{r cs-graph, eval=FALSE}
(g <- CTgraph(x))
plotCoordGraph(g)
```

The above representation greatly facilitates queries of the transformation(s) 
required to spatially align elements. `blobs_labels`, for example, requires a
sequential transformation (scaling and translation) for the `sequence` space:

```{r cs-path, results="hold", eval=FALSE}
# the following are equivalent
i <- "blobs_labels"
j <- "sequence"
y <- label(x, i)
CTpath(x, i, j)
invisible(CTpath(y, j))
```

## Plotting

```{r fig.width=8, fig.height=8}
sd_plot() + sd_plot_image(x) +
sd_plot() + sd_plot_point(x, c="x") +
sd_plot() + sd_plot_shape(x, 3) +
sd_plot() + sd_plot_label(x, c=c("lavender", "blue"))
```

```{r fig.width=4, fig.height=4}
sd_plot() + 
  coord_sf(reverse="y") +
  sd_plot_image(x) +
  sd_plot_label(x, c=rainbow(10), alpha=2/3) +
  sd_plot_shape(x, 3, fill="black", linewidth=1) +
  sd_plot_point(x, c="yellow", size=1/3)
```

# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
