## MCMICRO

In this demo, we'll analyze 12-plex CyCIF data on a human lung adenocarcinoma 
core, as output by MCMICRO [@Schapiro2022-MCMICRO], which stems from a larger 
tissue microarray (TMA); included are also cell segmentation outputs from both
*UnMICST* and *Ilastik*, respectively, as well as corresponding channel x cell 
protein expression matrices (here, corresponding to mean pixel intensities).

### preamble

```{r load-libs, message=FALSE, warning=FALSE}
library(sf)
library(dplyr)
library(ggplot2)
library(patchwork)
library(SpatialData)
library(SpatialData.data)
library(SpatialData.plot)
```

```{r load-data}
(sd <- SpatialData.data::LungAdenocarcinomaMCMICRO())
```

Let's start by viewing available `channels()`:

```{r}
cat(channels(image(sd)))
```

For consistency, we will fix a color to use for each channel:

```{r}
pal <- c(
    DNA_6="grey",
    DNA_7="grey",
    DNA_8="grey",
    CD57="maroon",  # NK
    FOXP3="yellow", # Treg
    SMA="red",      # stromal
    NCAM="orange",  # myocyte
    CD11B="blue",   # dendritic
    CD16="cyan",    # neutrophil
    ELANE="purple", # neutrophil
    ECAD="green",   # epithelial
    CD45="magenta") # lymphocyte
```

The original image includes a superfluous empty region, which we'll remove:

```{r}
sd <- crop(sd, list(xmin=0, xmax=2500, ymin=600, ymax=3200))
```

Now, let's generate a composite image of few selected channels:

```{r}
ch <- c("CD16", "CD45", "FOXP3", "ELANE", "ECAD", "SMA")
plotSpatialData() + plotImage(sd, ch=ch, c=pal[ch])
```

We can also visualize channels individually (here, using a smaller region):

```{r fig.height=6}
bb <- list(
    xmin=1600, xmax=2400, 
    ymin=1200, ymax=2000)
sp <- crop(sd, bb)

ch <- channels(image(sp))
ps <- lapply(ch, \(.) {
    plotSpatialData() + ggtitle(.) + 
    plotImage(sp, ch=., c=pal[.]) })

wrap_plots(ps, nrow=3) & 
    theme_void() & theme(
        legend.position="none", 
        plot.title=element_text(hjust=0.5))
```

```{r}
(se <- table(sd))
```

### appendix

::: {.callout-note icon=false, collapse=true}

### session

```{r sess-info}
sessionInfo()
```

:::
